SUPPORTING INFORMATION \\ Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy

### 2021-08-04 response_letter/response_letter.tex

\begin{figure}[ht!]
	\centering
    \includegraphics[angle=90,origin=c,width=1.0\textwidth]{../FigS5.pdf}
\end{figure}

\begin{figure}[ht!]
	\begin{center}
        \caption*{\textbf{Figure S5.} \rood{Macrostate interpretation and structural analysis. The Markov state model macrostates are learned from the simulation trajectories using the featurization and deep learning procedure described in Section 2.1.2 of the main text, but it is informative to perform a \textit{post hoc} projection of the simulation data into a low-dimensional space spanned by physically interpretable order parameters in order to determine the structural characteristics of each macrostate and the structural heterogeneity of the microstate ensemble. (a) Definition of two physically-interpretable order parameters useful in characterizing the structure and base pairing of each microstate under both shifting and fraying. We define the $3^\prime$ shift for a particular base on a strand as the linear distance between that base and the base on the complementary strand offset by two base pairs in the $3^\prime$ direction. This distance reaches a minimum when the two strands are mutually shifted two base pairs in the $3^\prime$ direction. We define $d_{3^\prime}$ as the mean $3^\prime$ shift averaged over the two central bases, represented by red arrows, in the complementary strands. It is more convenient to work the reciprocal $d_{3^\prime}^{-1}$, since it has a finite range whereas $d_{3^\prime}$ diverges for dehybridized strands. We define $d_{5^\prime}^{-1}$ analogously as the reciprocal mean $5^\prime$ shift, represented by green arrows. We have found this pair of variables extremely informative in revealing the microstate and macrostate structure and base pairing patterns. (b) We report the mean and standard deviation of $d_{3^\prime}$ and $d_{5^\prime}$ taken over all microstates comprising each macrostate for each sequence. The small standard deviations that are consistent with a narrow distribution of the bulk of the microstates around the free energy minimum of each macrostate. Cartoon renderings reflect the dominant configurations of each macrostate and how each is represented by the $d_{3^\prime}^{-1}$ and $d_{5^\prime}^{-1}$ order parameters green and red arrows, respectively. (c) Superposition of sample density for each macrostate for each sequence AT-all, GC-end, GC-core, and GC-mix projected into $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$. This panel conveys two important pieces of information. First, each macrostate is anchored around a single free energy minimum containing the preponderance of the probability mass of the microstate ensemble. These projections indicate a relatively narrowly peaked distribution of microstates around the free energy minimum with a broad tail. Second, the location of each macrostate free energy minimum in the $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$ projection readily illuminates the base pair pattern within the duplex. Within the fully hybridized H state (blue) $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 1.2 nm$^{-1}$. A 2 b.p.\ shift to 3$^\prime$ is manifest in a decrease in $d_{3^\prime}^{-1}$ (increase in $d_{3^\prime}$) and an increase in $d_{5^\prime}^{-1}$ (decrease in $d_{5^\prime}$) producing the stable 3S2 state (green). A further 2 b.p.\ shift to 3$^\prime$ results in a further decrease in $d_{3^\prime}^{-1}$ and also a decrease in $d_{5^\prime}^{-1}$ (back down to the same value as in the H state) to produce the 3S4 state (black). Analogous arguments apply to the 2 b.p.\ 5$^\prime$ shifted state 5S2 (orange) and 4 b.p.\ 5$^\prime$ shifted state 5S4 (purple). The dehybridized state D exists, due to the finite box size, at $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 0.2 nm$^{-1}$ (red). The GC-core frayed state F4 (orange) exists as a heart-shaped region between the H and D sates, with each symmetric lobe of the heart corresponding to fraying of one or other of the loose ends. (d) Coloring the microstates by their membership probabilities to their assigned macrostate under a fuzzy clustering procedure shows that microstates lying in the vicinity of the free energy minima have high probabilities whereas those at the macrostate cluster boundaries have less definitive assignments. The former comprise the preponderance of the microstate ensemble while the latter represent intermediate configurations that are transiently occupied as the system transitions from one macrostate to another.}}

@@

We guarded against overfitting using five-fold cross-validation in which we \rood{constructed five random partitions of the 40 independent simulation trajectories into a training set of 20 trajectories and a validation set of 20 trajectories}.\citep{Sidky2019High-ResolutionVAMPnets}

\textcolor{blue}{2)      For the validation of embedding dimensions (P. 11 L. 3), it is hard to judge the presence of a kink at 5D in the AT-all sequence since only a maximum of 6D is shown in Fig.~S1. Could a few additional dimensions be provided?}

\textbf{Author Reply:}   We agree that it is more difficult to see the kink for the AT-all sequence than the other sequences. As suggested by the reviewer, we added two additional modes to the cross-validation procedure that does indeed better expose this kink at five slow modes. We have updated the corresponding panel Fig.~S1 and also reproduce it below.

% Fig. S1
\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=0.8\textwidth]{../FigS1.pdf}
         \caption*{\textbf{Figure S1.} Five-fold cross-validation of the SRV VAMP-2 scores to select the optimal number of SRV coordinates for each sequence. A knee in the VAMP-2 plot was identified at the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively. An embedding of corresponding dimensionality was then used to cluster frames into discrete states. The absence of any significant separation in the training and testing VAMP-2 scores demonstrates that model is not overfitted.}
        \label{fig:SIFig1}
	\end{center}
\end{figure}

@@

\textcolor{blue}{3)      The values associated with the ``slow'' and ``fast'' rates appear to be inverse quantities, but this is not mentioned in the text (P. 14 L. 44).}

\textbf{Author Reply:}   The reviewer is correct that the rates have units s$^{-1}$ and describe the inverse of the experimental timescales shown on page 15. We agree that the sentence on is unclear in its current form and have made the following change to that text on p.~15:

``We conducted T-jump IR experiments for each DNA sequence as a function of temperature and extracted \rood{the ``slow'' $k_d^\mathrm{slow}$  and ``fast'' $k_d^\mathrm{fast}$ rates, corresponding to processes proceeding on 10-30 $\mu$s and 70-100 ns timescales, respectively.''}

We have also update the y-axis label in Fig.~1 to specify units in s$^{-1}$.

@@

\textcolor{blue}{4)      In Figure 1 the units for rates are not provided.}

\textbf{Author Reply:}   Thank you for catching this error. Fig.~1 has been updated to include units in s$^{-1}$.

\textcolor{blue}{5)      In the comparison between experiments and theory for slow and fast T-jump IR responses (P. 17 L. 9), were there any correlations between the goodness-of-fit in the calculated rates (Figure S4) and the agreement with experiments for a given system? It is hard to judge as is because the goodness-of-fit values were not provided with Figure S4.}

\textbf{Author Reply:}     We have updated all panels of Fig.~S4 plots to include $R^2$ values determined by linear least squares fits of the models in log-space (i.e., $\log \left( f \right) = -k_d t$) and have also re-plotted the data on log-linear axes for easier comparisons between the data and fits. The revised figure is reproduced below. In all cases except the lowest temperature T$_m$ - 5 K slow response fits where dissociation data is sparse, we report good fits with $R^2$ $>$ 0.88. We see marginally less good fits for the GC-core sequence compared to the other three systems, possibly -- as we observe in Section 3.1 of the main text -- due to its high propensity to fray, but the reduction in the quality of the fit is only marginal and in absolute terms the fits are still very good.

@@

\clearpage
\newpage

% Fig. S4
\begin{figure}[ht!]
	\centering
    \includegraphics[width=.75\textwidth]{../FigS4.pdf}
   \caption*{\textbf{Figure S4.} Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments. From 120 independent 1 $\mu$s simulations, we compiled the slow response data by recording the fraction of sequences with both central Watson-Crick base pairs intact as a function of time, and the fast response data as the fraction of sequences with both terminal Watson-Crick base pairs intact as a function of time. We define a Watson-Crick base pair to be intact if the centers-of-mass of the two comlementary bases lie within a linear distance of 1.3 nm. We extracted our computational estimate of $k_d^\mathrm{fast}$ by fitting a decaying exponential to the fraction of bound A:T termini as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$. Similarly, we extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$. \rood{We report within the figure legend of each panel the coefficient of determination $R^2$ for a least squares linear fit of the model to the data in log space (i.e., $\log \left( f \right) = -k_d t$) and data are plotted on log-linear axes to facilitate visual comparison of the fits. In all cases we observe excellent fits of the models to the data with all $R^2$ > 0.88 except for the slow response at the the lowest temperature T$_m$ - 5 K where dissociation events are sparse.}}
    \label{fig:SIFig4}
\end{figure}

@@

Second, the unbiased simulation data are in good agreement with trends in experimental T-jump IR measurements of fast and slow relaxation responses (Section 3.1) and are sufficient to furnish robust, low-variance, and well-converged MSMs (Section 3.2, Figs.~S1-3). This indicates that the data are of sufficient quality and volume to present favorable experimental comparisons and learn reliable long-time kinetic models.

Third, although it is straightforward to correct biased simulation data to recover the unbiased thermodynamics, it is more challenging to recover the unbiased kinetics. Although a number of approaches do exist, it has been our experience that the kinetic corrections can present difficulties in numerical convergence. That is not to say that we could not have successfully applied this methods, but just that we did not have cause to do so since the unbiased data were not unreasonably expensive to collect and showed favorable experimental comparisons and produced robust MSMs.

We did note the possibility of performing biased simulations in the Introduction on p.~4-5, and have now elaborated these remarks to better explain our choice to proceed with unbiased simulations:

``The effect of the applied bias upon the thermodynamics can be rigorously corrected for using standard reweighting techniques \citep{Gallicchio2005TemperaturePaths, Souaille2001ExtensionCalculations, Shirts2008StatisticallyStates, Kumar1992THEMethod}. \rood{Rigorous elimination of the bias in the kinetics is critical for the construction of robust and reliable kinetic models that reflect the true system dynamics and are uncontaminated by any residual effects of the biasing potentials used to induce good sampling. A number of approaches to correct the kinetics are also available, including Girsanov reweighting, transition-based reweighting analysis (TRAM), dynamic histogram analysis method (DHAM), and their derivatives~\citep{Prinz2011OptimalDynamics, Chodera2011DynamicalTemperatures, Stelzl2017DynamicSimulations, Donati2017GirsanovModels, Donati2018GirsanovSimulations, Quer2018ANCOORDINATES} \citep{Mey2014XTRAM:States}. The application of these methods under the conditions of high bias necessary for good sampling can, however, present challenges for numerical convergence.} A number of coarse-grained DNA force fields have been developed that enable direct observation of these events over microsecond time scales via unbiased coarse-grained molecular dynamics simulations, \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015, Dans2016MultiscaleDNA} which, up to an acceleration factor associated with the smoothing of the underlying free energy landscape inherent to the coarse-graining procedure, can preserve a faithful model of the unbiased dynamics and associated pathways. These models have previously been used to study biological phenomena such as nucleosome dynamics\citep{Lequieu2016Tension-dependentUnwrapping, Lequieu2017InSliding} and transcription factor binding \cite{Terakawa2015P53Simulations, Tan2018DynamicData} as well as nanotechnology applications such as strand displacement \citep{Srinivas2013OnDisplacement, Haley2020DesignDisplacement} and DNA origami \citep{Snodin2019Coarse-grainedOrigami, Doye2020TheOrigami}. \rood{In this study, we choose to employ a coarse-grained model for DNA \citep{Hinckley2013AnHybridization} that is sufficiently inexpensive to enable the collection of sufficient volumes of unbiased simulation trajectories and adequately sample configurational space that we do not need to appeal to biasing strategies to enhance convergence nor apply any \textit{post hoc} corrections to the thermodynamics or kinetics.}''

@@

\textcolor{blue}{The paper is generally well-written, and, I believe, represents a good attempt to tackle this important process. I personally have had a manuscript "desk-rejected" by JACS on the basis that it is "too focussed on DNA biotechnology", and "primarily of interest to RNA and DNA chemists". The same statements could be made about this paper. However, I disagree with the conclusion that such papers are not appropriate for JACS. I would therefore recommend publication, subject to satisfactory responses to the comments below.  }

\textbf{Author Reply:}   We are delighted by the reviewer's warm reception of our paper, and pleased that they find it to be appropriate for publication in JACS.

\textcolor{blue}{1.   I was frequently confused as to the temperatures used in the experiments and analysis. At one point, it is noted that simulations are performed at melting temperatures. However, simulations are actually performed at a range of temperatures and it isn't always clear when looking at a figure which temperatures were used for simulation, NN model comparison or experiments (see eg. Figs 2 and 3 and 5). I would recommend noting the temperature used to obtain the data in each case.}

\textbf{Author Reply:}   We thank the reviewer for this suggestion and regret our lack of clarity in this matter. All simulations performed in this work were conducted at the melting temperatures determined within the 3SPN.2 model, with the exception of the simulations discussed in Section 3.1 where we consider a range of temperatures in order to draw comparisons to experimental measurements over this temperature range. We have added a number of clarifications to the text in order to more clearly convey this important point to the reader.

@@

\clearpage
\newpage

\textcolor{blue}{6. Do the authors have any data on the scaling of k\_hybridization with T? In experiment, rates appear to increase with temperature (albeit with a much weaker dependence than de-hybridization rates). Simple "nucleation and zipper" models would predict the opposite, as does the oxDNA model (see ref 52).}

\textbf{Author Reply:}   This is an interesting question and we are, in fact, already considering embarking upon some additional work in this vein. Rigorous engagement of this question would require us to fit MSMs at a variety of temperatures in order to compare the pathways, mechanisms, and rates for hybridization and dehybridization. In the present work, we have only constructed MSMs at the melting temperature of each sequence and the collection of sufficient data to parameterize MSMs at other temperatures would require both (i) a substantial investment of time to collect the $\sim$1 ms of data required to fit robust MSMs and (ii) the use of biasing techniques in order to adequately sample hybridization (dehybridization) events at large temperature excursions above (below) the melting temperature. As such, we do not have the necessary data to engage this question at the present time.

@@

\begin{itemize}
\item $G_H$ = $G_D$ = 0 (adopting this as the arbitrary zero of free energy and asserting that the sequence is at its melting temperature so the hybridized H and dissociated D states are equiprobable)
\item $G_{F4}$ = 5.1 $kT$ at $T_m$ = 324 K = 13.7 kJ/mol
\item $f_H$ + $f_D$ + $f_F$ = 1 (the fractional populations sum to unity)
\item $f_X$ = $\exp (-\beta G_X$) / $Z$ ($Z$ is the partition function that imposes normalization)
\end{itemize}

We can solve this system of equations to yield $f_{F4}$ = 0.0030 that is in excellent agreement with the MSM predictions of $f_{F4}$ = 0.0028 that we report in Fig.~2b.

We have updated Figure 3 with the correct $\Delta$F as shown below.

\begin{figure}[ht!]
	\begin{center} 
        \includegraphics[width=140mm, scale=1]{../Fig3.pdf}
        \caption*{\textbf{Figure 3.} Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model \rood{at the sequence melting temperatures}.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} (a) Free energies of each macrostate relative to the hybridized state $\Delta F = F - F_H$. We define the hybridized state H to possess a free energy of zero and take care to only compare relative free energies (i.e., $\Delta F$) between the MSM and NN model. (b) Discrepancy between the macrostate relative free energy predictions $\Delta \Delta F = \Delta F^\mathrm{MSM} - \Delta F^\mathrm{NN}$ of the MSM relative to the NN model. The MSM tends to predict higher relative free energies (i.e., lower occupancy probabilities) relative to the hybridized state H compared to the NN model.}
        \label{fig:NN_table}
	\end{center}
\end{figure}

@@

\textbf{Author Reply:}   We have added these references to those at the top of p.~4.

\textcolor{blue}{12. The reasoning behind why the broader relaxation spectrum is indicative of misaligned states was not that clear to me. Why would "elevated" fraying cause a broader response? Wouldn't it also, then, be broader for CG-core, which has elevated fraying? It seems to me that the argument is more along the lines of the heterogeneity of the configurations leading to heterogeneous dynamics, and hence a broader relaxation profile. But I am no expert on this technique.}

\textbf{Author Reply:}   The reviewer makes a good comment on our choice of language used to described the stretched AT-all response. It is indeed our contention that the heterogeneity of the configurations in the initial ensemble and the fraying dynamics exhibited in each of these states that is the origin of the broadening of the relaxation profile, but our wording here was insufficiently clear. We believe that the reviewer's phrasing ``heterogeneity of configurations leads to heterogeneous dynamics'' expresses this idea much more clearly and we have integrated it into our edits of Section 3.7 on p.~33:

### 2021-08-04 response_letter/response_letter.tex

``Nevertheless, the presence of these out-of-register shifted states in the low-temperature equilibrium ensemble prior to the T-jump step may be observable via their influence on the fast T-jump IR response attributable to fraying. Specifically, we hypothesize that the dangling ends and inert tails present in the out-of-register shifted states should promote a \rood{broader fraying response over the course of the relaxation that is distinct from that of in-register fraying. This heterogeneity of configurations should lead to heterogeneous dynamics, manifested in the observation of a more stretched relaxation over experimental time scales of 70-100 ns.} Analysis of the MSM equilibrium distributions reveals 10.0\% of the equilibrium ensemble to reside in out-of-register shifted states 3S2, 5S2, 3S4, and 5S4 for AT-all, compared to just 0.23\% for GC-end, and 0\% for GC-core and GC-mix. It is our conjecture that a substantial population of out-of-register shifted states in the pre-T-jump AT-all ensemble should be distinguishable from the GC-end, GC-core, and GC-mix as an elongation of the fast relaxation response associated with terminal base fraying.''

@@

\textcolor{blue}{13. How much of the code is being made available? Ideally, the initialization files for the simulations and the code used to perform the MSM analysis would be publicly available. Additionally, it would be good to publish the raw experimental data and the processing tools used to obtain the final results.}

\textbf{Author Reply:}   We have uploaded input files and analysis scripts to Github (\url{https://github.com/mrjoness/Oligo-MSMs}) and 45 GB of featurized trajectories and experimental data to Zenodo (\url{https://doi.org/10.5281/zenodo.5148949}). The links to both of these resources are now noted in the Supporting Information statement on p.~38:

``\rood{Input files and analysis scripts have been uploaded to Github (\url{https://github.com/mrjoness/Oligo-MSMs}) and 45 GB of featurized trajectories and experimental data to Zenodo (\url{https://doi.org/10.5281/zenodo.5148949}).}''

@@

Updated.

\textcolor{blue}{4. The TOC graphic should fit in an area no larger than 3.25 in. by 1.75 in. (approx. 8.25 cm by 4.45  cm) and should have adequate resolution and clarity. Confirm that all text is legible at this size.}

Updated and confirmed.

============================================================

### 2021-08-04 SI.tex
JACS resubmission

%\newcommand*{\rood}[1]{{\color{red}{#1}}}
%\newcommand*{\rood}[1]{#1}
%\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
\newcommand*{\blauw}[1]{#1}
%\newcommand*{\groen}[1]{\textcolor{green}{#1}}
%\newcommand*{\groen}[1]{#1}
%\newcommand*{\blauwr}[1]{{\color{blue}{#1}}}
%\newcommand*{\blauwr}[1]{#1}

@@

% Fig. S4
\begin{figure}[ht!]
	\centering
    \includegraphics[width=.70\textwidth]{FigS4.pdf}
    \caption{Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments. From 120 independent 1 $\mu$s simulations, we compiled the slow response data by recording the fraction of sequences with both central Watson-Crick base pairs intact as a function of time, and the fast response data as the fraction of sequences with both terminal Watson-Crick base pairs intact as a function of time. We define a Watson-Crick base pair to be intact if the centers-of-mass of the two comlementary bases lie within a linear distance of 1.3 nm. We extracted our computational estimate of $k_d^\mathrm{fast}$ by fitting a decaying exponential to the fraction of bound A:T termini as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$. Similarly, we extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$.}
    \label{fig:SIFig4}
\end{figure}

@@

% Fig. S5
\begin{figure}[ht!]
	\centering
    \includegraphics[width=0.8\textwidth]{FigS5a.pdf}
\end{figure}
\begin{figure}[ht!]
	\centering
    \includegraphics[width=0.8\textwidth]{FigS5b.pdf}
\end{figure}

\begin{figure}[ht!]
	\centering
    \caption{GC-mix hybridizes by nucleation-zippering and dehybridizes by fraying-peeling. Tracking of the 10 intermolecular distances between native WC base pairs over the course of four additional (a) hybridization and (b) dehybridization events. Symmetrically permutable distances (e.g., A$_1$:T$_{10}$ and T$_{10}$:A$_1$) are reflected across the x-axis to avoid congestion in the plot. Circles superposed on the x-axis indicate the instantaneous MSM state assignment as dissociated D (blue) or hybridized H (orange). Hybridization tends to occur by a nucleation-zippering mechanism, wherein a native G:C pair and adjacent A:T pair or 2-3 central A:T pairs first form prior to rapid formation of the duplex. Dehybridization tends to occur by a fraying-peeling mechanism wherein fraying of the two-base AT-tails on one or both sides of the duplex precedes dissociation of the central native base pairs and complete dissolution of the duplex.}
    \label{fig:SIFig5}
\end{figure}

@@

The NN model assumes hybridization proceeds in a two-state all-or-nothing manner. The Gibbs free energy difference between the hybridized and dissociated states can be determined from calculation of NN enthalpy and entropy contributions,
\begin{equation}\label{SIe1}
\Delta G_{NN}^{\circ}(T) = \Delta H_{NN}^{\circ} - T\Delta S_{NN}^{\circ}.
\end{equation}
$\Delta H_{NN}^{\circ}$ and $\Delta S_{NN}^{\circ}$ are assumed to be independent of temperature $T$, and are computed from the sum over all NN terms in a given configuration,
\begin{equation}\label{SIe2}
\Delta H_{NN}^{\circ} = \sum_{i}^{n_{NN}}\Delta H_{NN, i}^{\circ} + \sum_{j}^{n_{DE}}\Delta H_{DE, j}^{\circ} + \Delta H_{init}^{\circ},
\end{equation}
\begin{equation}\label{SIe3}
\Delta S_{NN}^{\circ} = \sum_{i}^{n_{NN}}\Delta S_{NN, i}^{\circ} + \sum_{j}^{n_{DE}}\Delta S_{DE, j}^{\circ} + \Delta S_{init}^{\circ} + \Delta S_{sym}^{\circ}.
\end{equation}                          
$\Delta H_{NN, i}^{\circ}$ is the NN enthalpy for a given dinucleotide step within duplex DNA, $\Delta H_{DE, j}^{\circ}$ corresponds to the enthalpic contribution from a given dangling end (DE) base next to the duplex\citep{Santalucia2004TheMotifs}, and $\Delta H_{init}^{\circ}$ is a duplex initiation term that accounts for terminal effects on duplex enthalpy. A schematic illustration of the NN and DE contributions for a fully hybridized duplex and an out-of-register shifted state is presented in Fig.~\ref{fig:SIFig6}. $\Delta S_{NN}^{\circ}$ is constructed in an analogous manner from NN, dangling end, and initiation terms, plus a symmetry term, $\Delta S_{sym}^{\circ}$, that applies an entropic penalty to self-complementary sequences for maintaining C2 symmetry.

@@

@@

\begin{figure}[ht!]
	\centering
    \includegraphics[width=.85\textwidth]{FigS6.pdf}
    \caption{Schematic of nearest neighbor (NN) contributions for a fully hybridized AT-all duplex and an out-of-register duplex with dangling ends (DE).}
    \label{fig:SIFig6}
\end{figure}

>>>>

\newcommand*{\rood}[1]{{\color{red}{#1}}}
%\newcommand*{\rood}[1]{#1}
%\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
%\newcommand*{\blauw}[1]{#1}
%\newcommand*{\groen}[1]{\textcolor{green}{#1}}
%\newcommand*{\groen}[1]{#1}
%\newcommand*{\blauwr}[1]{{\color{blue}{#1}}}
%\newcommand*{\blauwr}[1]{#1}

@@

% Fig. S4
\begin{figure}[ht!]
	\centering
	\includegraphics[width=.97\textwidth]{FigS4.pdf}
\end{figure}

\begin{figure}[ht!]
	\begin{center}
    	\caption{Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments. From 120 independent 1 $\mu$s simulations, we compiled the slow response data by recording the fraction of sequences with both central Watson-Crick base pairs intact as a function of time, and the fast response data as the fraction of sequences with both terminal Watson-Crick base pairs intact as a function of time. We define a Watson-Crick base pair to be intact if the centers-of-mass of the two comlementary bases lie within a linear distance of 1.3 nm. We extracted our computational estimate of $k_d^\mathrm{fast}$ by fitting a decaying exponential to the fraction of bound A:T termini as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$. Similarly, we extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$. \rood{We report within the figure legend of each panel the coefficient of determination $R^2$ for a least squares linear fit of the model to the data in log space (i.e., $\log \left( f \right) = -k_d t$) and data are plotted on log-linear axes to facilitate visual comparison of the fits. In all cases we observe excellent fits of the models to the data with all $R^2$ > 0.88 except for the slow response at the the lowest temperature T$_m$ - 5 K where dissociation events are sparse.}}
    \label{fig:SIFig4}
    	\end{center}
\end{figure}

@@

\begin{figure}[ht!]
	\centering
	\includegraphics[angle=90,origin=c,width=1.0\textwidth]{FigS5.pdf}
\end{figure}

\begin{figure}[ht!]
	\begin{center}
        \caption{\rood{Macrostate interpretation and structural analysis. The Markov state model macrostates are learned from the simulation trajectories using the featurization and deep learning procedure described in Section 2.1.2 of the main text, but it is informative to perform a \textit{post hoc} projection of the simulation data into a low-dimensional space spanned by physically interpretable order parameters in order to determine the structural characteristics of each macrostate and the structural heterogeneity of the microstate ensemble. (a) Definition of two physically-interpretable order parameters useful in characterizing the structure and base pairing of each microstate under both shifting and fraying. We define the $3^\prime$ shift for a particular base on a strand as the linear distance between that base and the base on the complementary strand offset by two base pairs in the $3^\prime$ direction. This distance reaches a minimum when the two strands are mutually shifted two base pairs in the $3^\prime$ direction. We define $d_{3^\prime}$ as the mean $3^\prime$ shift averaged over the two central bases, represented by red arrows, in the complementary strands. It is more convenient to work the reciprocal $d_{3^\prime}^{-1}$, since it has a finite range whereas $d_{3^\prime}$ diverges for dehybridized strands. We define $d_{5^\prime}^{-1}$ analogously as the reciprocal mean $5^\prime$ shift, represented by green arrows. We have found this pair of variables extremely informative in revealing the microstate and macrostate structure and base pairing patterns. (b) We report the mean and standard deviation of $d_{3^\prime}$ and $d_{5^\prime}$ taken over all microstates comprising each macrostate for each sequence. The small standard deviations that are consistent with a narrow distribution of the bulk of the microstates around the free energy minimum of each macrostate. Cartoon renderings reflect the dominant configurations of each macrostate and how each is represented by the $d_{3^\prime}^{-1}$ and $d_{5^\prime}^{-1}$ order parameters green and red arrows, respectively. (c) Superposition of sample density for each macrostate for each sequence AT-all, GC-end, GC-core, and GC-mix projected into $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$. This panel conveys two important pieces of information. First, each macrostate is anchored around a single free energy minimum containing the preponderance of the probability mass of the microstate ensemble. These projections indicate a relatively narrowly peaked distribution of microstates around the free energy minimum with a broad tail. Second, the location of each macrostate free energy minimum in the $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$ projection readily illuminates the base pair pattern within the duplex. Within the fully hybridized H state (blue) $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 1.2 nm$^{-1}$. A 2 b.p.\ shift to 3$^\prime$ is manifest in a decrease in $d_{3^\prime}^{-1}$ (increase in $d_{3^\prime}$) and an increase in $d_{5^\prime}^{-1}$ (decrease in $d_{5^\prime}$) producing the stable 3S2 state (green). A further 2 b.p.\ shift to 3$^\prime$ results in a further decrease in $d_{3^\prime}^{-1}$ and also a decrease in $d_{5^\prime}^{-1}$ (back down to the same value as in the H state) to produce the 3S4 state (black). Analogous arguments apply to the 2 b.p.\ 5$^\prime$ shifted state 5S2 (orange) and 4 b.p.\ 5$^\prime$ shifted state 5S4 (purple). The dehybridized state D exists, due to the finite box size, at $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 0.2 nm$^{-1}$ (red). The GC-core frayed state F4 (orange) exists as a heart-shaped region between the H and D sates, with each symmetric lobe of the heart corresponding to fraying of one or other of the loose ends. (d) Coloring the microstates by their membership probabilities to their assigned macrostate under a fuzzy clustering procedure shows that microstates lying in the vicinity of the free energy minima have high probabilities whereas those at the macrostate cluster boundaries have less definitive assignments. The former comprise the preponderance of the microstate ensemble while the latter represent intermediate configurations that are transiently occupied as the system transitions from one macrostate to another.}}
        %\label{fig:macrostate_colors_FES}
	\end{center}
\end{figure}

@@

The NN model assumes hybridization proceeds in a two-state all-or-nothing manner. The Gibbs free energy difference between the hybridized and dissociated states can be determined from calculation of NN enthalpy and entropy contributions,
\begin{equation}\label{SIe1}
\Delta G_{NN}^{\circ}(T) = \Delta H_{NN}^{\circ} - T\Delta S_{NN}^{\circ}.
\end{equation}
$\Delta H_{NN}^{\circ}$ and $\Delta S_{NN}^{\circ}$ are assumed to be independent of temperature $T$, and are computed from the sum over all NN terms in a given configuration,
\begin{equation}\label{SIe2}
\Delta H_{NN}^{\circ} = \sum_{i}^{n_{NN}}\Delta H_{NN, i}^{\circ} + \sum_{j}^{n_{DE}}\Delta H_{DE, j}^{\circ} + \Delta H_{init}^{\circ},
\end{equation}
\begin{equation}\label{SIe3}
\Delta S_{NN}^{\circ} = \sum_{i}^{n_{NN}}\Delta S_{NN, i}^{\circ} + \sum_{j}^{n_{DE}}\Delta S_{DE, j}^{\circ} + \Delta S_{init}^{\circ} + \Delta S_{sym}^{\circ}.
\end{equation}                          
$\Delta H_{NN, i}^{\circ}$ is the NN enthalpy for a given dinucleotide step within duplex DNA, $\Delta H_{DE, j}^{\circ}$ corresponds to the enthalpic contribution from a given dangling end (DE) base next to the duplex\citep{Santalucia2004TheMotifs}, and $\Delta H_{init}^{\circ}$ is a duplex initiation term that accounts for terminal effects on duplex enthalpy. A schematic illustration of the NN and DE contributions for a fully hybridized duplex and an out-of-register shifted state is presented in Fig.~\ref{fig:SIFig7}. $\Delta S_{NN}^{\circ}$ is constructed in an analogous manner from NN, dangling end, and initiation terms, plus a symmetry term, $\Delta S_{sym}^{\circ}$, that applies an entropic penalty to self-complementary sequences for maintaining C2 symmetry.

@@

\rood{In applying the NN model to each macrostate, we made the simplifying assumption that the ensemble of microstates constituting each macrostate could be represented by a single pattern of Watson-Crick base pairing that are schematically illustrated in Fig.~2a. In reality, each macrostate is composed of an ensemble of microstates that may have slightly different base pairing patterns and explicitly averaging over all of these microstates may change the predictions of the NN model. Recalculating the NN predictions by explicitly averaging over all microstates in the macrostate leads to changes in the $\Delta F = F - F_H$ values reported in Fig.~3 (i.e., the free energy of each macrostate relative to the hybridized state) of $<$1 kJ/mol. As such, representing each macrostate by a single representative microstate and base pairing pattern is a highly accurate simplifying approximation for the purposes of this calculation.}

@@

\begin{figure}[ht!]
	\centering
    \includegraphics[width=.85\textwidth]{FigS6.pdf}
    \caption{Schematic of nearest neighbor (NN) contributions for a fully hybridized AT-all duplex and an out-of-register duplex with dangling ends (DE).}
    \label{fig:SIFig7}
\end{figure}

% Fig. S7
\begin{figure}[ht!]
	\centering
    \includegraphics[width=0.8\textwidth]{FigS7a.pdf}
\end{figure}
\begin{figure}[ht!]
	\centering
    \includegraphics[width=0.8\textwidth]{FigS7b.pdf}
\end{figure}

\begin{figure}[ht!]
	\centering
    \caption{GC-mix hybridizes by nucleation-zippering and dehybridizes by fraying-peeling. Tracking of the 10 intermolecular distances between native WC base pairs over the course of four additional (a) hybridization and (b) dehybridization events. Symmetrically permutable distances (e.g., A$_1$:T$_{10}$ and T$_{10}$:A$_1$) are reflected across the x-axis to avoid congestion in the plot. Circles superposed on the x-axis indicate the instantaneous MSM state assignment as dissociated D (blue) or hybridized H (orange). Hybridization tends to occur by a nucleation-zippering mechanism, wherein a native G:C pair and adjacent A:T pair or 2-3 central A:T pairs first form prior to rapid formation of the duplex. Dehybridization tends to occur by a fraying-peeling mechanism wherein fraying of the two-base AT-tails on one or both sides of the duplex precedes dissociation of the central native base pairs and complete dissolution of the duplex.}
    \label{fig:SIFig6}
\end{figure}
